---
title: "Final Project Description"
author: "Matt Higham"
output: 
  rmdformats::readthedown:
    code_folding: show
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Collaboration:

* the final project is an __individual__ project. Unlike the previous projects, you may __not__ collaborate with other students. This project is a chance for you to show what you've learned throughout the semester.
* you may use the Internet and StackExchange, but you also should not copy- paste code directly from the website, without citing that you did so.
* the `reprex` system worked well for Project 2 so any code questions that you have must be put in to a `reprex`. Though you’ll have to take a few minutes to put your question into a `reprex`, the nice thing about this method of answering questions is you’ll get a direct answer to your question instead of a "hint."
* Part 2 will ask for an `R` code appendix. You can easily add in all of your `R` code to the end of your .html file by putting this chunk at the end of your .Rmd file (and removing the hashtags):

```{r, echo = TRUE, eval = FALSE}
## ```{r printcode, all-code, ref.label=knitr::all_labels(),  echo=TRUE, eval=FALSE}
## ```
```

__Due Date__: __Friday, December 11 at 12:00 pm EST__.

__Turn in your final project as a zip file that contains all files for your Project, including one or two .Rmd files, one or two .html files, the data set that you used, and any other materials necessary to run your Project code. Parts 1 and 2 can be in separate .Rmd files or in the same .Rmd file.__

## Part 1: General Exploration (100 points)

For Part 1, you will do an exploratory analysis of your chosen data set (a _hypothesis-exploration_ analysis). Choose a few specific questions that you want to answer and then answer those questions with graphs and tables. A few notes:

* there's no need for your questions to be related at all to Part 2, but some of them could be related. For example, for a sports project, you might be interested in looking at how accurate the win probabilities seem to be, or, for the music project, you might explore relationships that don't have any direct impact on predicting the artist.
* for scope, use Mini-Projects 1 and 2 as guides: I'd expect the analysis to be of similar detail (for example, Mini-Project 2 answered 3 broad questions/topics, looking at scores over time, difference in scores between courses, and strengths and weaknesses compared to the university average.)
* for the sports/babynames projects, you should __not__ include static graphics of your animation in this section, as those will be required in Part 2.

<br>

__Points for Part 1:__

* Introduction (10 points). The data set should be introduced, including an introduction of the variables of interest that will be used throughout the report. A quick preview of the questions of interest that will be answered throughout the report should also be given.
* Content (30 points). Are the questions that will be answered interesting and of an appropriate level for this class? You should aim for an overall level that’s about the same as the questions given in Mini-Project 2, which had 3 overarching questions that were each answered with a table, a graphic, or both, coupled with a written explanation. Aim for a rough length and level of Mini-Project 2, without the "on your own" exercise at the end.
* Correctness (25 points). Does the report (tables, graphs, and text) actually answer the questions posed with appropriate graphics and tables, or, if the questions cannot be answered, give a reasonable explanation for why?
* Conclusion and Summary (10 points). Consider any limitations in your analysis. These could be limitations of the data that is used or limitations in your analysis directly. Also, provide a brief summary of your findings here.
* Formatting (15 points). This is our usual "does the report look and sound nice and professional" criterion.
* Appendix (10 points). As an Appendix, include a broad explanation of what your strategy and thought process was for this project. How did you decide which questions you wanted to answer? What packages did you need to use? Were there any major stumbling blocks? etc. 

<br>

## Part 2: knn or `gganimate` (100 points)

In Part 2, you'll dive into your specific project topic. 

__For knn, points will be allocated by:__

* Introduction (5 points). You can re-use some of what you've written in Part I, but should now explain the overall goal of your knn model as well.
* Methods (10 points). Present which predictors you chose to include in your final model, as well as results from your __training model__. Include any explanations of scaling that you did and/or any subsetting of the data that you needed to perform.
* Results (15 points). Present your confusion matrix from the __test/holdout sample__ and discuss any patterns that you see. Additionally, present the proportion of times your model correctly predicted the artist.
* Discussion (20 points). Run the `new_artists.csv` data (found on Sakai in the project folder) through your knn algorithm (making sure to do any necessary scaling and transformation) and incorporate the following into a discussion: (a) Why __can't__ you make a confusion matrix using this new data, (b) What do your results tell you, (c) How might this relate to Spotify's "Suggested for You" algorithm, and (d) Are you surprised by any of the results from this? Why or why not?
* Summary and Wrap-Up (5 points) Provide a 1-2 sentence wrap-up and, if you were to work on this further, what would you look into or improve upon?

<br>

* Formatting (10 points) This is our usual "does the report look and sound nice and professional" criterion.
* Appendix: `R` code (10 points) Include an Appendix with all of your `R` code (you can include code from Part 1 or leave it out). The 10 points here are allocated to how "readable" your code is and if the code is correct.
* Appendix: knn Understanding (15 points) Provide an explanation of knn in about two paragraphs. Include things like what the algorithm is used for, why it's useful, what $k$ means, why you need to scale variables, and anything else you find useful from any readings.
* Appendix: Validation Understanding (10 points) Explain the process for how you chose your predictors in your model and for how you chose $k$. Include things like how you measured what a "good" model was and how and why you split your original data set in two.

<br>

__For `gganimate`, points will be allocated by:__

* Introduction (5 points). You can re-use some of what you've written in Part I, but should now explain the overall goal of your animation as well.
* Introduction: ELO Explanation (5 points) Write a paragraph that explains what ELO is, as well as what it's strengths and weaknesses are. For `babynames`, do a little research on the SSA and where the `babynames` data comes from <https://www.ssa.gov/oact/babynames/limits.html>. 
* Initial Explorations (10 points). Provide a few static graphics of what your animation is for a few different points in time. You likely made these before adding the code for the animation. Comment on any trends that you see. 

<br>

* Result 1 (20 points). Provide the animation of the top $n$ teams or top $n$ baby names through time, similar to the example graphic. You can choose $n$ to be Top 5, 10, 15, etc. The animation should look as nice and professional as the example.
* Discussion 1 (5 points). Write a brief paragraph that explains what your animation shows.
* Result 2 (15 points). Provide a different animation through time. For example, instead of looking at ELO through years, you might consider looking at ELO weekly for a single season. For the `babynames`, I can't think of any other obvious graphics to make, as only year is given. You can think of something yourself, or, as an alternative, you can make a static graphic with time on the x-axis and number of babies on the y-axis that shows a trend in traditionally "boy" names becoming more popular "girl" names and vice-versa.
* Discussion 2 (5 points). Write a brief paragraph that explains what your animation (or, for `babynames`, static graphic) shows.
* Summary and Wrap-Up (5 points) Provide a 1-2 sentence wrap-up and, if you were to work on this further, what would you look into or improve upon?

<br>

* Formatting (10 points) This is our usual "does the report look and sound nice and professional" criterion.
* Appendix: `R` code (10 points) Include an Appendix with all of your `R` code (you can include code from Part 1 or leave it out). The 10 points here are allocated to how "readable" your code is and if the code is correct.
* Appendix: Data Wrangling Explanation (10 points) Provide an overview of what "pre-processing" steps you needed to perform to prepare the data set for the animation plot. 

<br>

__Turn in your final project as a zip file that contains all files for your Project, including one or two .Rmd files, one or two .html files, the data set that you used, and any other materials necessary to run your Project code. Parts 1 and 2 can be in separate .Rmd files or in the same .Rmd file.__

<br>
