---
title: "Week 2: Data Wrangling with `dplyr`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

Suppose we have the following toy data set, named `df`. The first two columns are numeric while the third column is categorical.

```{r}
library(tidyverse)
df <- tibble(x1 = c(1, 7, 4),
  x2 = c(3, NA, 2),
  cat1 = c("Yes", "Yes", "No"))
library(pander)
pander(df)
```

__Create New Variables with `mutate()` (Perhaps with `case_when()` or `if_else()`)__

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

__Choose Rows to Keep with `filter()`__

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

__Choose Columns to Keep with `select()`__

\pagebreak

```{r}
pander(df)
```

__Order/Sort Your Data Set with `arrange()`__

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

__Obtain Numerical Summaries with `summarise()`__

&nbsp;

&nbsp;

&nbsp;

&nbsp;

__Obtain Numerical Summaries by Group with `group_by()` and `summarise()`__



