---
title: "9: Strings with `stringr` Handout"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = "hide")
```

Suppose we have the following data set with a couple of variables containing strings. Our example is small, but you might think of each variable as containing lyrics to a song or the text of a book or essay.

```{r, results = "as.is"}
library(tidyverse)
library(pander)
df1 <- tibble(strings_var1 = c("abc", "hi there"), strings_var2 = c("def", "okay byeeee"))
pander(df1)
```

The following are just a few functions from the `stringr` package to manipulate strings.

`str_c()` to combine two or more strings.

```{r}
str_c(df1$strings_var1, df1$strings_var2)
```

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

`str_split()` to Split a String in Two or More Parts

```{r}
str_split(df1$strings_var1, pattern = " ", simplify = TRUE)
```

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

`str_remove()` to Remove a Pattern from a String

```{r}
str_remove(df1$strings_var2, pattern = "eee")
```

&nbsp;

&nbsp;

&nbsp;

\pagebreak

`str_sub()` to grab certain parts of a string.

```{r}
str_sub(df1$strings_var1, start = 1, end = 2)
## similar to separate()
```

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

`str_detect()` to Detect whether a String has a Certain Pattern

```{r}
str_detect(df1$strings_var2, pattern = "e")
```

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

Using `^` and `$` in `str_detect()` 

```{r}
str_detect(df1$strings_var2, pattern = "^e")
str_detect(df1$strings_var2, pattern = "e$")
```

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

* You can also use `.` to match any character, `\d` to match any digit, `\s` to match any whitespace, `[abc]` to match a, b, or c, and `[^abc]` to match anything except a, b, or c.


