---
title: "Midterm 1: Biases in Course Evaluations?"
output: 
  rmdformats::readthedown:
    code_folding: hide
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.keep = "none", message = FALSE, warning = FALSE)
```

There are __165__ total points for this exam. Good luck!

__Note__:  You can leave tables as is and graphs without labels, titles, without changing colours, etc, unless specifically noted in a problem.

__Note 2__: Problems are roughly organized from easy to difficult so it's recommended to go through them from start to finish (so that you definitely attempt all of the easier problems). Additionally, each problem is completely independent, in that you'll never need something from an earlier problem to complete a later problem.

__Note 3__: Two or three questions ask you to interpret the graph or table. Please don't forget to do this!

__Note 4__: The approximate time is given next to each question. This times assume that the exam is __90__ minutes, so, if you stick with the approximate times throughout the exam, you'll still have an extra __45__ minutes at the end.

Create a new .Rmd file. Then, copy and paste the following statement and type in your name below:

"I affirm that all work on this exam is entirely my own." - 

<br>

<br>

The `openintro` package has a data set called `evals` that is described by the `openintro` authors as:

"The data are gathered from end of semester student evaluations for a large sample of professors from the University of Texas at Austin. In addition, six students rate the professors' physical appearance. The result is a data frame where each row contains a different course and each column has information on either the course or the professor."

In the data set we're working with, `evals_tidy`, each professor has 2 courses for which they are being rated on, and each row corresponds to a course.

Read in the data with

```{r, echo = TRUE, class.source = "fold-show"}
library(tidyverse)
evals_tidy <- read_csv("evals_tidy.csv")
head(evals_tidy)
```

Variables that we will use include:

* `prof_id`, in identification number assigned to each professor in the study.
* `score`, the average evaluation score for that course (out of 5)
* `rank` of the professor, a categorical variable that is either `tenured`, `tenure track`, or `teaching`.
* `ethnicity` of the professor, either `minority` or `not minority`
* `gender` of the professor, either `female` or `male`
* `age` of the professor (years)
* `cls_students`, the number of registered students in the course
* `bty_avg`, the attractiveness of the professor, as rated by a separate set of students not in the professor's course
